Exploring probability distributions for bivariate temporal granularities

### 2019-12-02 main.Rmd
data structure

Effective exploration or good visualization require good data structures. Temporal data problems are typically grouped into two types of analysis, time series and longitudinal. A new tidy data structure to support exploration and modeling of temporal data is tsibble, where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. Commonly, simple sequences of time value pairs $<t_0, v_0> \dots, <t_n, v_n>$ are the basis of analysis and visualization. It is recognized that the initial approaches of just considering time as an ordinal dimension in a 2D or 3D visualisation are inadequate to capture characteristics of time-dependent information. For example, there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use different calendars that define meaningful system of granularities for different application domain. Data can then be integrated at different levels of granularity enabling value aggregation or distribution along granularities. Since, we are essentially concerned with detection of previously unknown periodic behavior of data, it makes sense to support the detection of patterns by anabling aggregation or distribution across cyclic time granularities. We consider the following data structure for exploring distribution of a measured variable ((Figure \@ref(fig:data-structure)). These are based on conditioning the values taken on by one or more of the cyclic granularities in a data set. This means data subsets are formed for each level of the conditioned variable and the fixed variable. This can be an useful data structure for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

```{r data-structure, fig.cap = "The data structure for exploring periodicities along cyclic granularities. Two cyclic granularities serve as fixed and conditioned variable and the observational unit across combinations of these cyclic time granularities.", fig.pos="htb"}
include_graphics("Figs/data-structure.png")
```

>>>>

Effective exploration or good visualization require good data structures. Commonly, simple sequences of time value pairs $<t_0, v_0> \dots, <t_n, v_n>$ are the basis of analysis and visualization. It is recognized that the initial approaches of just considering time as an ordinal dimension in a 2D or 3D visualisation are inadequate to capture characteristics of time-dependent information. There is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis.  Moreover, one can use calendars based on application domain that define meaningful system of granularities. Data can then be consolidated for different levels of granularity enabling statistical summaries of values along granularities. Since, we are interested in detection of previously unknown periodic behavior of data, it makes sense to support the detection of patterns by obtaining statistical summaries across cyclic time granularities. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, the big problem can be broken down into smaller pieces by focusing on two cyclic granularities at a time.

A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the bottom granularity, which is the index of the tsibble, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. Data subsets are formed for each level of the conditioned variable and the fixed variable. This can be an useful data structure for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

```{r data-structure, fig.cap = "The data structure for exploring periodicities in data. Two cyclic granularities serve as fixed and conditioned variable and the measurement variable is the observation across combinations of these cyclic time granularities.", fig.pos="htb"}
include_graphics("Figs/data-structure-v1.png")
```
